---
title: "SpaTalk Advanced Usage"
author: 
  - name: "Zaoqu Liu"
    email: "liuzaoqu@163.com"
    affiliation: "Maintainer"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{SpaTalk Advanced Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This vignette covers advanced usage scenarios for SpaTalk, including:

- Custom ligand-receptor databases
- Alternative deconvolution methods
- Parallel processing optimization
- Working with different ST platforms
- Troubleshooting common issues

## Custom Databases

### Custom Ligand-Receptor Pairs

You can use your own curated LR pairs instead of the built-in CellTalkDB:

```{r custom_lr, eval=FALSE}
# Custom LR pairs must have these columns:
# - ligand: ligand gene symbol
# - receptor: receptor gene symbol  
# - species: "Human" or "Mouse"

custom_lrpairs <- data.frame(
  ligand = c("CXCL12", "CCL2", "VEGFA"),
  receptor = c("CXCR4", "CCR2", "KDR"),
  species = "Human"
)

# Use in find_lr_path
obj <- find_lr_path(obj, lrpairs = custom_lrpairs, pathways = pathways)
```

### Custom Pathway Database

```{r custom_pathway, eval=FALSE}
# Custom pathways must have:
# - src: source gene
# - dest: destination gene (downstream)
# - type: interaction type (optional)

custom_pathways <- data.frame(
  src = c("CXCR4", "CXCR4", "STAT3"),
  dest = c("STAT3", "AKT1", "MYC"),
  type = c("activation", "activation", "transcription")
)

obj <- find_lr_path(obj, lrpairs = lrpairs, pathways = custom_pathways)
```

## Alternative Deconvolution Methods

SpaTalk supports multiple deconvolution backends:

### Method 1: Built-in NNLM (Default)

```{r nnlm, eval=FALSE}
obj <- dec_celltype(
  object = obj,
  sc_data = sc_data,
  sc_celltype = sc_celltype,
  method = 1  # NNLM
)
```

### Method 2: RCTD (spacexr)

```{r rctd, eval=FALSE}
# Requires: install.packages("spacexr")
obj <- dec_celltype(
  object = obj,
  sc_data = sc_data,
  sc_celltype = sc_celltype,
  method = 2  # RCTD
)
```

### Method 3: Seurat Integration

```{r seurat, eval=FALSE}
obj <- dec_celltype(
  object = obj,
  sc_data = sc_data,
  sc_celltype = sc_celltype,
  method = 3  # Seurat
)
```

### Method 4: SPOTlight

```{r spotlight, eval=FALSE}
# Requires: BiocManager::install("SPOTlight")
obj <- dec_celltype(
  object = obj,
  sc_data = sc_data,
  sc_celltype = sc_celltype,
  method = 4  # SPOTlight
)
```

### Method 5: deconvSeq

```{r deconvseq, eval=FALSE}
# Requires: devtools::install_github("reneshbedre/deconvSeq")
obj <- dec_celltype(
  object = obj,
  sc_data = sc_data,
  sc_celltype = sc_celltype,
  method = 5  # deconvSeq
)
```

### Method 6: stereoscope (Python)

```{r stereoscope, eval=FALSE}
# Requires: conda environment with stereoscope
obj <- dec_celltype(
  object = obj,
  sc_data = sc_data,
  sc_celltype = sc_celltype,
  method = 6,  # stereoscope
  python_path = "/path/to/conda/envs/stereoscope/bin/python"
)
```

### Method 7: cell2location (Python)

```{r cell2location, eval=FALSE}
# Requires: conda environment with cell2location
obj <- dec_celltype(
  object = obj,
  sc_data = sc_data,
  sc_celltype = sc_celltype,
  method = 7,  # cell2location
  python_path = "/path/to/conda/envs/cell2location/bin/python"
)
```

## Parallel Processing

### Enabling Parallel Processing

SpaTalk leverages `doParallel` for multi-core computation:

```{r parallel, eval=FALSE}
library(doParallel)

# Detect available cores
n_cores <- parallel::detectCores() - 1
cat("Using", n_cores, "cores\n")

# Register parallel backend
registerDoParallel(cores = n_cores)

# Run with parallel enabled
obj <- dec_cci_all(
  object = obj,
  if_doParallel = TRUE
)

# Clean up
stopImplicitCluster()
```

### Memory Optimization

For large datasets, consider:

```{r memory, eval=FALSE}
# Process cell types in batches
celltypes <- unique(obj@meta$rawmeta$celltype)
batch_size <- 5

for(i in seq(1, length(celltypes), batch_size)) {
  batch <- celltypes[i:min(i + batch_size - 1, length(celltypes))]
  
  for(ct_sender in batch) {
    for(ct_receiver in batch) {
      obj <- dec_cci(obj, ct_sender, ct_receiver)
    }
  }
  
  gc()  # Force garbage collection
}
```

## Platform-Specific Workflows

### 10x Visium

```{r visium, eval=FALSE}
# Load Visium data (via Seurat)
library(Seurat)
visium <- Load10X_Spatial("/path/to/visium/")

# Extract data for SpaTalk
st_data <- GetAssayData(visium, slot = "counts")
st_meta <- data.frame(
  spot = colnames(visium),
  x = visium@images$slice1@coordinates$col,
  y = visium@images$slice1@coordinates$row
)

# Create SpaTalk object (spot-based)
obj <- createSpaTalk(
  st_data = st_data,
  st_meta = st_meta,
  species = "Human",
  if_st_is_sc = FALSE,
  spot_max_cell = 10  # Expected cells per spot
)
```

### Slide-seq

```{r slideseq, eval=FALSE}
# Slide-seq typically has ~10 cells per bead
obj <- createSpaTalk(
  st_data = slideseq_counts,
  st_meta = slideseq_coords,
  species = "Mouse",
  if_st_is_sc = FALSE,
  spot_max_cell = 10
)
```

### STARmap / MERFISH (Single-cell resolution)

```{r starmap, eval=FALSE}
# Single-cell resolution - no deconvolution needed
obj <- createSpaTalk(
  st_data = starmap_counts,
  st_meta = starmap_coords,
  species = "Mouse",
  if_st_is_sc = TRUE,
  spot_max_cell = 1,
  celltype = cell_annotations  # Pre-annotated cell types
)
```

## Extracting Results

### LR Pair Results

```{r extract_lr, eval=FALSE}
# Get significant LR pairs
lr_results <- obj@lrpair
sig_pairs <- lr_results[lr_results$lr_co_ratio_pvalue < 0.05, ]

# Export to CSV
write.csv(sig_pairs, "significant_lr_pairs.csv", row.names = FALSE)
```
### Downstream TF Scores

```{r extract_tf, eval=FALSE}
# Get TF activity scores
tf_results <- obj@tf

# Filter by score threshold
active_tfs <- tf_results[tf_results$score > 0.1, ]
```

### Full CCI Network

```{r extract_network, eval=FALSE}
# Combine LR and TF results into network
network <- merge(
  obj@lrpair,
  obj@tf,
  by = c("celltype_sender", "celltype_receiver")
)

# Export for Cytoscape
write.csv(network, "cci_network.csv", row.names = FALSE)
```

## Troubleshooting

### Common Issues

**Issue: "Gene not found in ST data"**
```{r gene_not_found, eval=FALSE}
# Check gene overlap
st_genes <- rownames(obj@data$rawdata)
lr_genes <- unique(c(lrpairs$ligand, lrpairs$receptor))
overlap <- intersect(st_genes, lr_genes)
cat("Overlapping genes:", length(overlap), "\n")
```

**Issue: Memory errors with large datasets**
```{r memory_error, eval=FALSE}
# Reduce data size
obj <- createSpaTalk(
  st_data = st_data[, sample(ncol(st_data), 5000)],  # Subsample spots
  ...
)
```

**Issue: No significant LR pairs found**
```{r no_sig, eval=FALSE}
# Check expression thresholds
# Lower the expression cutoff
obj <- find_lr_path(
  obj, lrpairs, pathways,
  min_exp = 0.01  # Lower threshold
)
```

## Best Practices

1. **Data Quality Control**
   - Filter low-quality spots/cells before analysis
   - Normalize expression data appropriately

2. **Reference Selection**
   - Use tissue-matched scRNA-seq reference
   - Ensure cell type annotations are accurate

3. **Parameter Tuning**
   - Adjust `spot_max_cell` based on platform
   - Use appropriate distance thresholds for tissue type

4. **Validation**
   - Cross-validate with known biological interactions
   - Compare results across replicates

## Session Info

```{r session}
sessionInfo()
```
