---
title: "SpaTalk: Knowledge-Graph-Based Cell-Cell Communication Inference"
author: 
  - name: "Zaoqu Liu"
    email: "liuzaoqu@163.com"
    affiliation: "Maintainer"
  - name: "Xin Shao"
    email: "xin_shao@zju.edu.cn"
    affiliation: "Original Developer"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{SpaTalk: Knowledge-Graph-Based Cell-Cell Communication Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

**SpaTalk** is a computational framework for inferring spatially resolved cell-cell communications (CCIs) from spatial transcriptomics (ST) data. Published in *Nature Communications* (2022), SpaTalk integrates graph network modeling and knowledge graph approaches to reconstruct ligand-receptor-target signaling networks between spatially proximal cells.

### Citation

> Shao, X., Li, C., Yang, H., et al. Knowledge-graph-based cell-cell communication inference for spatially resolved transcriptomic data with SpaTalk. *Nature Communications* **13**, 4429 (2022). https://doi.org/10.1038/s41467-022-32111-8

### Key Features

- **Cell-type deconvolution** for spot-based ST data using NNLM
- **Spatial mapping** between scRNA-seq and ST data
- **Knowledge graph-based** ligand-receptor-target pathway inference
- **Permutation-based** statistical validation
- Support for **single-cell** and **spot-based** ST platforms

## Installation

```{r install, eval=FALSE}
# From R-universe (recommended)
install.packages("SpaTalk", repos = "https://zaoqu-liu.r-universe.dev")

# From GitHub
devtools::install_github("Zaoqu-Liu/SpaTalk")
```

## Quick Start

### Load Package and Data

```{r load}
library(SpaTalk)

# Load built-in demo data
load(system.file("extdata", "starmap_data.rda", package = "SpaTalk"))
load(system.file("extdata", "starmap_meta.rda", package = "SpaTalk"))

# Load curated databases
data(lrpairs)
data(pathways)

cat("Expression matrix:", nrow(starmap_data), "genes x", ncol(starmap_data), "cells\n")
cat("Metadata:", nrow(starmap_meta), "cells\n")
cat("LR pairs:", nrow(lrpairs), "\n")
```

### Create SpaTalk Object

```{r create}
# Prepare spatial metadata
st_meta <- data.frame(
  cell = starmap_meta$cell,
  x = starmap_meta$x,
  y = starmap_meta$y
)

# Create SpaTalk object (single-cell resolution data)
obj <- createSpaTalk(
  st_data = starmap_data,
  st_meta = st_meta,
  species = "Mouse",
  if_st_is_sc = TRUE,
  spot_max_cell = 1,
  celltype = starmap_meta$celltype
)

obj
```

### Filter LR-Pathway Pairs

```{r filter}
# Find LR pairs with downstream pathway targets
obj <- find_lr_path(
  object = obj,
  lrpairs = lrpairs,
  pathways = pathways,
  if_doParallel = FALSE
)

cat("Filtered LR pairs:", nrow(obj@lr_path$lrpairs), "\n")
```

### Infer Cell-Cell Communications

```{r cci, eval=FALSE}
# Infer CCIs between specific cell types
obj <- dec_cci(
  object = obj,
  celltype_sender = "eL6",
  celltype_receiver = "PVALB"
)

# Or infer all pairwise CCIs
obj <- dec_cci_all(object = obj, if_doParallel = FALSE)
```

## Session Info

```{r session}
sessionInfo()
```
