---
title: "SpaTalk Visualization Guide"
author: 
  - name: "Zaoqu Liu"
    email: "liuzaoqu@163.com"
    affiliation: "Maintainer"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{SpaTalk Visualization Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)
```

## Introduction

SpaTalk provides a comprehensive suite of visualization functions for exploring spatial transcriptomics data and cell-cell communication results. This vignette demonstrates all available plotting functions with examples.

## Setup

```{r load}
library(SpaTalk)
library(ggplot2)

# Load demo data
load(system.file("extdata", "starmap_data.rda", package = "SpaTalk"))
load(system.file("extdata", "starmap_meta.rda", package = "SpaTalk"))
data(lrpairs)
data(pathways)

# Create SpaTalk object
st_meta <- data.frame(
  cell = starmap_meta$cell,
  x = starmap_meta$x,
  y = starmap_meta$y
)

obj <- createSpaTalk(
  st_data = starmap_data,
  st_meta = st_meta,
  species = "Mouse",
  if_st_is_sc = TRUE,
  spot_max_cell = 1,
  celltype = starmap_meta$celltype
)

obj <- find_lr_path(obj, lrpairs, pathways, if_doParallel = FALSE)
```

## Spatial Cell Type Visualization

### plot_st_celltype

Visualize cell type distributions in spatial coordinates.

```{r celltype, eval=FALSE, fig.cap="Spatial distribution of cell types"}
plot_st_celltype(
  object = obj,
  celltype = unique(starmap_meta$celltype)[1:3],
  size = 1.5
)
```

### plot_st_celltype_all

Display all cell types in a single spatial plot.

```{r celltype_all, eval=FALSE, fig.cap="All cell types in spatial context"}
plot_st_celltype_all(
  object = obj,
  size = 1
)
```

### plot_st_celltype_percent

Show cell type composition as pie charts per spot (for spot-based data).
```{r celltype_percent, eval=FALSE, fig.cap="Cell type proportions per spot"}
# For spot-based ST data after deconvolution
plot_st_celltype_percent(
  object = obj_spot,
  celltype = c("Macrophage", "Fibroblast", "T_cell")
)
```

### plot_st_celltype_density

Kernel density estimation of cell type spatial distributions.

```{r celltype_density, eval=FALSE, fig.cap="Cell type density map"}
plot_st_celltype_density(
  object = obj,
  celltype = "eL6"
)
```

## Gene Expression Visualization

### plot_st_gene

Visualize gene expression patterns in spatial coordinates.

```{r gene, eval=FALSE, fig.cap="Spatial gene expression"}
# Get available genes
genes <- rownames(obj@data$rawdata)[1:2]

plot_st_gene(
  object = obj,
  gene = genes[1],
  size = 1.5
)
```

## Cell-Cell Communication Visualization

After running `dec_cci()` or `dec_cci_all()`, various CCI visualizations become available.

### plot_cci_lrpairs

Chord diagram showing ligand-receptor interactions between cell types.

```{r cci_chord, eval=FALSE, fig.cap="LR pair interactions (chord diagram)"}
plot_cci_lrpairs(
  object = obj,
  celltype_sender = "eL6",
  celltype_receiver = "PVALB"
)
```

### plot_lrpair

Spatial visualization of specific ligand-receptor pair interactions.

```{r lrpair, eval=FALSE, fig.cap="Spatial LR pair visualization"}
plot_lrpair(
  object = obj,
  ligand = "Cxcl12",
  receptor = "Cxcr4",
  celltype_sender = "Fibroblast",
  celltype_receiver = "T_cell"
)
```

### plot_lrpair_vln

Violin plot comparing ligand/receptor expression across cell types.

```{r lrpair_vln, eval=FALSE, fig.cap="LR expression violin plot"}
plot_lrpair_vln(
  object = obj,
  ligand = "Cxcl12",
  receptor = "Cxcr4"
)
```

## Pathway Visualization

### plot_lr_path

Network visualization of ligand-receptor-pathway connections.

```{r lr_path, eval=FALSE, fig.cap="LR-pathway network"}
plot_lr_path(
  object = obj,
  ligand = "Cxcl12",
  receptor = "Cxcr4",
  celltype_sender = "Fibroblast",
  celltype_receiver = "T_cell"
)
```

### plot_path2gene

Heatmap of pathway gene expression.

```{r path2gene, eval=FALSE, fig.cap="Pathway-gene heatmap"}
plot_path2gene(
  object = obj,
  celltype_receiver = "T_cell",
  receptor = "Cxcr4"
)
```

## Correlation Analysis

### plot_st_cor_heatmap

Correlation heatmap between cell types or gene sets.

```{r cor_heatmap, eval=FALSE, fig.cap="Cell type correlation heatmap"}
plot_st_cor_heatmap(
  object = obj,
  celltype = c("eL6", "PVALB", "Astro", "SST")
)
```

## Cell-Cell Distance Distribution

### plot_ccdist

Distribution of distances between interacting cell types.

```{r ccdist, eval=FALSE, fig.cap="Cell-cell distance distribution"}
plot_ccdist(
  object = obj,
  celltype_sender = "eL6",
  celltype_receiver = "PVALB"
)
```

## Customization Tips

### Color Palettes

SpaTalk uses ggplot2 for all visualizations. Customize colors easily:

```{r colors, eval=FALSE, fig.cap="Custom color palette example"}
library(ggplot2)

# Create base plot
p <- plot_st_celltype_all(obj, size = 1)

# Add custom theme
p + 
  scale_color_brewer(palette = "Set2") +
  theme_minimal() +
  labs(title = "Custom Styled Plot")
```

### Combining Plots

Use patchwork or cowplot for multi-panel figures:

```{r combine, eval=FALSE}
library(patchwork)

p1 <- plot_st_celltype(obj, celltype = "eL6")
p2 <- plot_st_celltype(obj, celltype = "PVALB")

p1 + p2 + plot_layout(ncol = 2)
```

## Summary of Visualization Functions

| Function | Description | Input Required |
|----------|-------------|----------------|
| `plot_st_celltype` | Single cell type spatial | SpaTalk object |
| `plot_st_celltype_all` | All cell types spatial | SpaTalk object |
| `plot_st_celltype_percent` | Pie chart per spot | Deconvolved object |
| `plot_st_celltype_density` | Density heatmap | SpaTalk object |
| `plot_st_gene` | Gene expression spatial | SpaTalk object |
| `plot_st_pie` | Pie chart composition | Deconvolved object |
| `plot_cci_lrpairs` | CCI chord diagram | After dec_cci |
| `plot_lrpair` | LR pair spatial | After dec_cci |
| `plot_lrpair_vln` | LR violin plot | After dec_cci |
| `plot_lr_path` | LR-pathway network | After dec_cci |
| `plot_path2gene` | Pathway heatmap | After dec_cci |
| `plot_st_cor_heatmap` | Correlation heatmap | SpaTalk object |
| `plot_ccdist` | Distance distribution | SpaTalk object |

## Session Info

```{r session}
sessionInfo()
```
