---
title: "SpaTalk Algorithm: Methodological Framework"
author: 
  - name: "Zaoqu Liu"
    email: "liuzaoqu@163.com"
    affiliation: "Maintainer"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{SpaTalk Algorithm: Methodological Framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

## Overview

SpaTalk employs a multi-step computational framework to infer spatially resolved cell-cell communications from spatial transcriptomics data. This vignette provides detailed explanations of the underlying algorithms.
## Algorithmic Pipeline

```
┌─────────────────────────────────────────────────────────────────┐
│                     SpaTalk Pipeline                            │
├─────────────────────────────────────────────────────────────────┤
│  1. Cell-type Deconvolution (for spot-based ST)                 │
│     └── Non-negative Linear Model (NNLM)                        │
│                                                                 │
│  2. Spatial Reconstruction                                      │
│     └── Monte Carlo sampling + k-NN spatial mapping             │
│                                                                 │
│  3. LR-Pathway Filtering                                        │
│     └── Knowledge graph: LR pairs → Downstream TFs              │
│                                                                 │
│  4. CCI Inference                                               │
│     └── Co-expression analysis + Permutation test               │
│                                                                 │
│  5. Pathway Scoring                                             │
│     └── Random walk on gene-gene interaction graph              │
└─────────────────────────────────────────────────────────────────┘
```

## 1. Cell-Type Deconvolution

### Mathematical Formulation

For spot-based ST data where each spot contains multiple cells, SpaTalk decomposes the observed expression profile into cell-type-specific contributions using **Non-negative Matrix Factorization (NMF)**.

Given:
- $X \in \mathbb{R}^{G \times S}$: ST expression matrix (genes × spots)
- $R \in \mathbb{R}^{G \times C}$: Reference expression profile (genes × cell types)

The deconvolution problem is formulated as:

$$X \approx R \cdot W$$

where $W \in \mathbb{R}^{C \times S}$ represents cell-type proportions per spot, subject to:
- $W_{ij} \geq 0$ (non-negativity)
- $\sum_i W_{ij} = 1$ (proportions sum to 1)

### NNLM Algorithm

SpaTalk uses the NNLM (Non-Negative Linear Model) package with the following optimization:

$$\min_{W \geq 0} \|X - RW\|_F^2 + \lambda \|W\|_1$$

The L1 regularization term promotes sparsity in cell-type assignments.

```{r nnlm_demo, eval=FALSE}
# Deconvolution workflow
obj <- dec_celltype(
  object = obj,
  sc_data = sc_counts,      # scRNA-seq reference
  sc_celltype = cell_labels, # Cell type annotations
  if_doParallel = TRUE
)
```

## 2. Spatial Reconstruction

### Monte Carlo Spatial Sampling

After deconvolution, SpaTalk reconstructs single-cell resolution expression by sampling cells from the reference data according to:

1. **Cell proportion sampling**: For each spot $s$ with predicted proportions $W_s$, sample $N_s$ cells where $N_s$ is the expected number of cells per spot.

2. **Spatial coordinate assignment**: Assign coordinates to sampled cells using a weighted distance model:

$$P(x_i | s) \propto \exp\left(-\frac{d(x_i, c_s)^2}{2\sigma^2}\right)$$

where $c_s$ is the spot centroid and $\sigma$ controls the spatial spread.

### k-Nearest Neighbor Mapping

For each reconstructed cell, find the k-nearest neighbors in the original ST data to refine expression estimates:

$$\hat{E}_i = \frac{1}{|N_k(i)|} \sum_{j \in N_k(i)} E_j \cdot w_{ij}$$

where $w_{ij}$ is the distance-based weight.

## 3. Knowledge Graph Integration

### LR-Pathway Database

SpaTalk integrates curated biological knowledge from multiple databases:

| Database | Content | Size |
|----------|---------|------|
| CellTalkDB | Ligand-receptor pairs | 3,398 (Human), 2,033 (Mouse) |
| KEGG | Signaling pathways | ~300 pathways |
| Reactome | Pathway interactions | ~2,000 pathways |
| AnimalTFDB | Transcription factors | ~1,600 TFs |

### Pathway Filtering Algorithm

```{r pathway_filter, eval=FALSE}
# The find_lr_path function filters LR pairs with downstream targets
obj <- find_lr_path(obj, lrpairs, pathways)

# Filter criteria:
# 1. Both ligand and receptor expressed in ST data
# 2. Downstream pathway genes present
# 3. Target TFs expressed
```

## 4. Cell-Cell Communication Inference

### Co-expression Score

For a ligand-receptor pair (L, R) between sender cell type A and receiver cell type B:

$$\text{CoExp}(L_A, R_B) = \frac{|\{(a,b) : E_L^a > 0 \land E_R^b > 0 \land d(a,b) < \delta\}|}{|\{(a,b) : d(a,b) < \delta\}|}$$

where $\delta$ is the spatial distance threshold.

### Permutation Test

Statistical significance is assessed via permutation:

1. Permute cell type labels $n$ times (default: 1000)
2. Recalculate co-expression for each permutation
3. Compute p-value:

$$p = \frac{|\{\pi : \text{CoExp}^\pi \geq \text{CoExp}^{obs}\}| + 1}{n + 1}$$

```{r permutation, eval=FALSE}
# CCI inference with permutation testing
obj <- dec_cci(
  object = obj,
  celltype_sender = "Macrophage",
  celltype_receiver = "Fibroblast",
  n_permutation = 1000
)
```

## 5. Downstream Pathway Scoring

### Random Walk Algorithm

To score downstream transcription factors activated by a receptor, SpaTalk performs random walks on the gene-gene interaction (GGI) graph:

**Algorithm: Random Walk TF Scoring**

```
Input: GGI graph G, receptor R, TF set T, n_walks, max_hop
Output: TF scores

Initialize: score[tf] = 0 for all tf in T

for i = 1 to n_walks:
    current = R
    for hop = 1 to max_hop:
        neighbors = G.neighbors(current)
        if neighbors is empty: break
        next = random_choice(neighbors)
        if next in T:
            score[next] += 1
        current = next

Return: score / n_walks
```

### C++ Implementation

SpaTalk uses optimized C++ code for performance-critical computations:

```cpp
// Simplified random walk implementation
NumericVector cpp_random_walk(
    CharacterVector ggi_src,
    CharacterVector ggi_dest,
    String receptor_name,
    CharacterVector tf_names,
    int n_walks = 10000,
    int max_hop = 10
) {
    // Build adjacency list
    // Perform random walks
    // Count TF visits
    // Return normalized scores
}
```

## Computational Complexity

| Step | Time Complexity | Space Complexity |
|------|-----------------|------------------|
| Deconvolution | $O(G \cdot S \cdot C)$ | $O(G \cdot S)$ |
| Spatial reconstruction | $O(S \cdot N \cdot k)$ | $O(N)$ |
| Co-expression | $O(L \cdot N^2)$ | $O(N^2)$ |
| Permutation test | $O(P \cdot L \cdot N^2)$ | $O(N^2)$ |
| Random walk | $O(W \cdot H)$ | $O(|E|)$ |

Where: G = genes, S = spots, C = cell types, N = cells, L = LR pairs, P = permutations, W = walks, H = max hops, E = edges.

## References

1. Shao, X., et al. (2022). Knowledge-graph-based cell-cell communication inference for spatially resolved transcriptomic data with SpaTalk. *Nature Communications*, 13, 4429.

2. Lin, X., & Boutros, P. C. (2020). Optimization and expansion of non-negative matrix factorization. *BMC Bioinformatics*, 21, 7.

3. Kanehisa, M., & Goto, S. (2000). KEGG: Kyoto Encyclopedia of Genes and Genomes. *Nucleic Acids Research*, 28(1), 27-30.

## Session Info

```{r session}
sessionInfo()
```
